---
title: "Rotate text"
author: "Steve Simon"
date: "October 31, 2016"
output: html_document
---

There are times when you want to put text on a diagonal slope, rather than vertical or horizontal. Diagonals are tricky because you need to adjust between the slope as measured in terms of the x-y coordinates and the slope as measured in terms of the physical plot itself, typically in pixels. Here's an example where you might want diagonal text.

```{r rotation_1}
plot(0, 0, xlim=c(0, 20), ylim=c(0, 100), type="n", xlab=" ", ylab=" ")
x0 <-  5
x1 <- 10
x2 <- 15
y0 <- 20
y1 <- 50
y2 <- 80
lines(x=c(x0, x2, x2, x0), y=c(y0, y0, y2, y0))
text(x1, y0-1.5*strheight("a", units="user"), "a")
text(x2+1.5*strwidth("b", units="user"), y1, "b")
delta.x <- x2 - x0
delta.y <- y2 - y0
ctext <- expression(sqrt(a^2+b^2))
text(x1, y1+1.5*strheight(ctext, units="user"), ctext)
```

That looks nice, but wouldn't it be better if you could tilt the formula on the hypotenuse? You can do this with the srt argument, but you need transform the dimension in the x-y coordinates to the dimension of inches or pixels.

```{r rotation_2}
plot(0, 0, xlim=c(0, 20), ylim=c(0, 100), type="n", xlab=" ", ylab=" ")
lines(x=c(x0, x2, x2, x0), y=c(y0, y0, y2, y0))
text(x1, y0-1.5*strheight("a", units="user"), "a")
text(x2+1.5*strwidth("b", units="user"), y1, "b")
px.per.xy <- par("cra") / par("cxy")
hyp.angle <- atan2(delta.y*px.per.xy[2], delta.x*px.per.xy[1]) * 180 / pi
text(x1, y1+1.5*strheight(ctext, units="user"), ctext, srt=hyp.angle)
```

Let's look at the conversion in detail. If you calculated the slope in x-y coordinates, you would get an angle of `r round(atan2(delta.y, delta.x) * 180 / pi)` degrees. That's because in x-y coordinates, the line is very steep. The horizontal distance is only `r delta.x` units, but the vertical distance is `r delta.y` units. But when you look at the graph on the screen, the horizontal distance looks to be almost twice as big as the vertical distance. That's because there are `r round(px.per.xy[1], 1)` pixels for each unit in the x direction and `r round(px.per.xy[2], 1)` pixels for each unit in the y direction. That means that you have `r round(delta.x*px.per.xy[1])` horizontal pixels and `r round(delta.y*px.per.xy[2])` vertical pixels. The arctangent in pixels will give you an angle of `r round(hyp.angle)` degrees.